---
title: "Reproducible Research: Peer Assessment 1"
author: "Paul Hartzog"
output: html_document
keep_md: true
---

###Loading and preprocessing the data

Read in the .csv file for the dataset and make sure date is set as.Date
```{r load data, results=FALSE}
setwd("E:/Users/Paul/Desktop/Reproducible research/Course Project 1/repdata_data_activity/")
data<-read.csv("activity.csv")
data$date <-as.Date(data$date)
```

###What is mean total number of steps taken per day?

Histogram of the total number of steps taken each day. Load the plyr package to use ddply.
```{r histogram}
library(plyr)
tot.steps<-ddply(data, .(date), summarise, total=sum(steps, na.rm=TRUE))
hist(tot.steps$total, main="Histogram of Total Steps", xlab="Total Steps")
```

Mean and median number of steps each day.
```{r means}
ddply(data, .(date), summarise, mean=mean(steps, na.rm=TRUE))
ddply(data, .(date), summarise, median=median(steps, na.rm=TRUE))
```

###What is the average daily activity pattern?

Daily activity pattern. 
```{r 5 min intervals}
inter<-ddply(data, .(interval), summarise, mean=mean(steps, na.rm=TRUE))
plot(inter$interval, inter$mean, type="l", main="Mean Number of Steps for Each Interval", ylab="Mean", xlab="Interval")
```

Interval with the average max number of steps
```{r max}
inter[which.max(inter$mean),1]
```

###Inputting missing values

Calculation of the total number of values missing from the data
```{r total}
data2<-data[!is.na(data$steps),]
nrow(data)-nrow(data2)
```
Creating new dataset with the missing values filled. I used 0 because it was the median for every day.
```{r fill}
data.nona<-data
data.nona$steps[is.na(data.nona$steps)]<-0
```
Recreating the histogram with missing values filled
```{r full hist}
tot.steps.nn<-ddply(data.nona, .(date), summarise, total=sum(steps, na.rm=TRUE))
hist(tot.steps.nn$total, main="Histogram of Total Steps", xlab="Total Steps")
```

Recalculate the mean and median for each day and compare to orginal with NAs removed
```{r full mean med}
means_old<-ddply(data, .(date), summarise, mean=mean(steps, na.rm=TRUE))
medians_old<-ddply(data, .(date), summarise, median=median(steps, na.rm=TRUE))
means_nn<-ddply(data.nona, .(date), summarise, mean=mean(steps, na.rm=TRUE))
medians_nn<-ddply(data.nona, .(date), summarise, median=median(steps, na.rm=TRUE))
means_nn<-cbind(means_nn, means_old[,2])
colnames(means_nn)<-c("date", "mean", "old mean")
medians_nn<-cbind(medians_nn, medians_old[,2])
colnames(medians_nn)<-c("date", "median", "old median")
means_nn
medians_nn
```
***Making all NAs equal to zero made some of the dates that previously had no data have mean and median of 0***

###Are there differences in activity patterns between weekdays and weekends?

First create a new column with the day of the week.
```{r weekdays}
data.nona$day<-weekdays(data$date)
```
Change each to a weekday or weekend
```{r weekday or weekdn}
data.nona$day<-gsub("Monday","Weekday", data.nona$day)
data.nona$day<-gsub("Tuesday","Weekday", data.nona$day)
data.nona$day<-gsub("Wednesday","Weekday", data.nona$day)
data.nona$day<-gsub("Thursday","Weekday", data.nona$day)
data.nona$day<-gsub("Friday","Weekday", data.nona$day)
data.nona$day<-gsub("Saturday","Weekend", data.nona$day)
data.nona$day<-gsub("Sunday","Weekend", data.nona$day)
```
Load lattice and create the panel plot that compares weekdays and weekends
```{r weekday plot}
library(lattice)
inter.nona<-ddply(data.nona, .(interval, day), summarise, mean=mean(steps, na.rm=TRUE))
xyplot(mean ~ interval|day, data=inter.nona, type="l", xlab="Interval", ylab="Mean Number of Steps")
```
